\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-01-18 paper.tex
add missing steps in MLMM
technical docu for the MLMM class

\section*{Software}
All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \url{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.

@@

\paragraph*{E-step.}
Supposing that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$, we need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters, which is given by 
\[\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].\]
Here, computing this quantity reduces to computing $\E_{\theta^{(w)}}[b_i | y_i]$ and $\E_{\theta^{(w)}}[b_ib_i^T | y_i]$ for $i = 1,\ldots,n$.
The marginal distributions of $y_i$ and $b_i$ being both Gaussian, one has from Bayes Theorem
\[ f(b_i|y_i; \theta) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i)^\top \Omega_i^{-1} (b_i - \mu_i) \big\}\]
where
\[\Omega_i = ({V_i}^\top \Sigma_iV_i + D^{-1})^{-1} \quad \text{ and } \quad \mu_i = \Omega_i{V_i}^\top \Sigma_i(y_i - U_i\beta).\]
Then, on has
\begin{center}
  $\left\{
    \begin{aligned}
      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i, \\
      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i + \mu_i\mu_i^\top.
    \end{aligned}
    \right.$
\end{center}

@@

>>>>

\section*{Software}
All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \url{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial. A technical documentation is also provided in Appendix~\ref{sec:package}.

@@

\paragraph*{E-step.}
Supposing that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$, we need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters, which is given by 
\[\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].\]
Here, computing this quantity reduces to computing $\E_{\theta^{(w)}}[b_i | y_i]$ and $\E_{\theta^{(w)}}[b_ib_i^T | y_i]$ for $i = 1,\ldots,n$.
The marginal distributions of $y_i$ and $b_i$ being both Gaussian, one has from Bayes Theorem
\[ f(b_i|y_i; \theta^{(w)}) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i^{(w)})^\top {\Omega_i^{(w)}}^{-1} (b_i - \mu_i^{(w)}) \big\}\]
where
\[\Omega_i^{(w)} = ({V_i}^\top \Sigma_i^{(w)}V_i + {D^{(w)}}^{-1})^{-1} \quad \text{ and } \quad \mu_i^{(w)} = \Omega_i^{(w)}{V_i}^\top \Sigma_i^{(w)}(y_i - U_i\beta^{(w)}).\]
Then, on has
\begin{center}
  $\left\{
    \begin{aligned}
      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i^{(w)}, \\
      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i^{(w)} + \mu_i^{(w)}{\mu_i^{(w)}}^\top.
    \end{aligned}
    \right.$
\end{center}

@@

\section{Technical documentation of the \texttt{lights} package}
\label{sec:package}

\subsection{\texttt{MLMM class}}

This class implements an EM algorithm for fitting a multivariate linear mixed model used to initialize the MCQNEM algorithm.
Let us introduce the list $\Omega^{(w)} = [\Omega_1^{(w)}, \ldots, \Omega_n^{(w)}]$, the matrices $\mu = [\mu_1, \ldots, \mu_n] \in \R^{r \times n}$, $U^l = [{U_{1l}}^\top \cdots {U_{nl}}^\top]^\top \in \R^{n_l \times q_l}$, $U = [U_1^\top \cdots U_n^\top]^\top \in \R^{\cN \times q}$,
\[ V^l = \begin{bmatrix}
    V_{1l} & \cdots & 0\\
    \vdots &  \ddots & \vdots \\
    0 & \cdots & V_{nl}
\end{bmatrix},
\quad
V = \begin{bmatrix}
  V_{1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{n}
\end{bmatrix}
\quad \text{ and } \quad
{\Omega^l}^{(w)} = \begin{bmatrix}
    {\Omega_{1}^l}^{(w)} & \cdots & 0\\
    \vdots &  \ddots & \vdots \\
    0 & \cdots & {\Omega_{n}^l}^{(w)}
\end{bmatrix}\]
that belong respectively in $\R^{n_l \times nr_l}$, $\R^{\cN \times nr}$ and $\R^{nr_l \times nr_l}$, as well as the vectors $\tilde \mu^{(w)} = ({\mu_1^{(w)}}^\top \cdots {\mu_n^\top)^{(w)}}^\top \in \R^{nr}$, $(\tilde \mu^l)^{(w)} = \big({(\mu_1^l)^{(w)}}^\top \cdots {(\mu_n^l)^{(w)}}^\top \big)^\top \in \R^{nr_l}$, $y^l = ({y_1^l}^\top  \cdots {y_n^l}^\top)^\top \in \R^{n_l}$
with $n_l = \sum_{i=1}^{n} n_i^l$ and $y = (y_1^\top \cdots y_n^\top)^\top \in \R^\cN$ with $\cN = \sum_{i=1}^{n} n_i$.
The $\beta$ update then rewrites
\[\beta^{(w+1)} = ({U}^\top U)^{-1}{U}^\top(y-V \tilde \mu^{(w)}).\]
For the $D$ update, one has
\[D^{(w+1)} = n^{-1} \big( \text{sum}(\Omega^{(w)}) + \mu^{(w)}{\mu^{(w)}}^\top\big).\]
And finally for the $\phi$ update, one has
\begin{align*}
\phi_l^{(w+1)} = n_l^{-1} \big[&(y^l - U^l{\beta_l^{(w+1)})}^\top \big(y^l - U^l\beta_l^{(w+1)} - 2 V^l (\tilde\mu^l)^{(w)}\big) \\
&+ \text{Tr}\big\{{V^l}^\top V^l \big({\Omega^l}^{(w)} + (\tilde\mu^l)^{(w)} {(\tilde\mu^l)^{(w)}}^\top\big)\big\} \big].
\end{align*}
